Emittance preservation of an electron bunch in a loaded quasi-linear plasma wakefield

### 2017-08-30 Tex/Paper.tex
More text tinkering

\vspace{1em}
\noindent\textbf{From Patric, expand and add this:}\\
Then I would propose to state that the SM does not create a focusing force increasing linearly with radius because not
all e- are evacuated. The state that because the structure is small, a  bunch with significant charge and reasonable
current is not much shorter than the structure size and thus large energy spread is expected.

@@

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors that
may affect this, we have tried several approaches to create a stable drive beam structure based on previous self-
modulation studies.

@@

In order to evaluate the quality of the beam, we also needed to study emittance evolution. However, full PIC codes like
Osiris are vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect'' \cite{godfrey:1974}.
This is a know issue with the Yee EMF solver which causes the phase velocity of electromagnetic fields to be lower than
$c$ while the beam moves very close to $c$. The effect can be mitigated somewhat by replacing the Yee solver with a
solver designed by Lehe \cite{lehe:2013}. Our simulations showed that this numerical effect is still prominent in the
high density regions of the electron beam, making it difficult to distinguish emittance growth from the physics from
that originating from numerical error.

Quasi-static PIC codes do not suffer this problem, so in order to study the emittance evolution of the beam we instead
turned to the recently released open source version QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D PIC
code \cite{huang:2006, an:2013}.

@@

In the QuickPIC simulations we use a single proton drive bunch that sets up a wakefield comparable to that which we
expect to see from the self-modulated SPS beam in AWAKE Run 2. The baseline AWAKE drive beam contains $3\nexp{11}$
protons \cite{gschwendtner:2016}. Only half of these are driving the wakefield, presuming the self-modulation is seeded
in the middle of the unmodulated bunch. In addition. a significant portion of the protons are lost during
self-modulation. The needed gap between the two plasma cells also contributes to a decreased density of protons near the
axis, and some of the field is drained by other protons seeing an accelerating phase within the drive beam itself. In
total, the expected peak $E_{z}$ field of $~2\funit{GV}{m}$ in the first plasma cell drops to $500-600\funit{MV}{m}$ in
the second cell \cite{awake_collaboration:2016}. A single proton bunch of $1.46\nexp{10}$, or $2.34\unit{nC}$ and
$7\unit{kA}$ was able to drive an $E_{z}$ field of a comparable magnitude in our single drive bunch setup.

The single bunch approach also has the benefit of reducing simulation time for the 3D simulations. Our aim was to create
a stable environment for the witness beam. We are not considering the evolution of the proton beam in this test setup,
so to prevent the proton beam from evolving radially, we increased the drive beam particle mass by a factor of
$1\mexp{6}$.

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. The
plasma electrons are depleted to around $65\%$ of nominal plasma density at the injection point of the electron beam in
reference simulations \cite{awake_collaboration:2016}. This condition is replicated in out single bunch case. The peak
density of the drive beam is $0.83\cdot n_{0}$ producing a quasi-linear condition behind it. The single bunch setup uses
the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size $\sigma_{r} = 200\unit{\mu m}$
throughout the plasma cell. The length of the drive bunch $\sigma_{z} = 40\unit{\mu m}$.

@@

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. AWAKE baseline energy for the long witness beam for Run 1 is $W_{eb} = 10-20\unit{MeV}$
\cite{gschwendtner:2016}. In Run 2 the experiment will likely require a higher initial energy, but a
$W_{eb} \approx 50\unit{MeV}$ is sufficient to minimise de-phasing as the drift $\Delta\xi \propto \gamma^{-2}$
\cite{berglyd_olsen:2015}. For this case we have eliminated this source of de-phasing entirely in order to reduce the
number of variables.

Earlier simulations with witness beam radius $\sigma_{r} = 100\unit{\mu m}$ caused the beam to undergo rapid transverse
evolution as the beam entered the plasma region, reducing its $\sigma_{r}$ from $100\unit{\mu m}$ to $< 10\unit{\mu m}$
\cite{berglyd_olsen:2016}. This was primarily due to a mismatch between beam emittance and the plasma density. This
relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength.

In these simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{r}$ of $5.25\unit{\mu m}$. The relatively small size of the witness beam compared to the proton beam put some
restrictions on the transverse resolution of the witness beam. The transverse grid cell resolution was set to
$1.17\unit{\mu m}$.

>>>>

With this experimental design the self-modulated proton beam does not produce a non-linear wakefield, and not all plasma
electrons are therefore evacuated. The result is that the focusing force will not increase linearly with radius. The
small size of the accelerating structure produced by the self-modulated beam also put constraints on the size of the
electron witness bunch, and thus its charge and current, if we wish to prevent large energy spread. By matching the
transverse size to a practical emittance and the plasma density, we prevent large amplitude oscillations which may cause
additional growth in energy spread as well as in emittance.

This beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. For a reasonable initial emittance of $2\unit{\mu m}$, this corresponds to
a $\sigma_{r}$ of $5.25\unit{\mu m}$ -- a very narrow bunch compared to the drive beam initial
$\sigma_{r} = 200\unit{\mu m}$. This charge density of this compact witness bunch reaches that of the plasma density at
only a few $\unit{pC}$. It only and reaches optimal beam loading at around $1-200\unit{pC}$, the implication being that
the witness bunch produces its own non-linear wakefield already at the head of the bunch. The majority of the electrons
within it will therefore see a linear focusing force preventing the bulk of the beam from undergoing emittance growth.

@@

The main focus of this study has been on the beam loading of the electron bunch. In order to eliminate other factors
that may affect this, we have tried several approaches to create a stable drive beam structure based on previous self-
modulation studies.

@@

In order to evaluate the quality of the accelerated bunch it was also necessary to study emittance evolution. However,
full PIC codes like Osiris are vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect''
\cite{godfrey:1974}. This is a know issue with the Yee EMF solver which causes the phase velocity of electromagnetic
fields to be lower than $c$ while the beam moves very close to $c$. The effect can be mitigated somewhat by replacing
the Yee solver with a solver designed by Lehe \cite{lehe:2013}. Our simulations showed that this numerical effect is
still prominent in the high density regions of our very compact electron bunch, making it difficult to distinguish
emittance growth from the physics from that originating from numerical error, and for long simulations drowning it
entirely.

In order to study the witness bunch emittance evolution we turned to the recently released open source version of
QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D quasi-static PIC code \cite{huang:2006, an:2013}. The
numerical Cherenkov effect does not apply to such quasi-static codes.

@@

In these QuickPIC simulations we use a single proton drive bunch that sets up a wakefield comparable to that which we
expect to see from the self-modulated SPS beam in AWAKE Run 2. The baseline AWAKE drive beam contains $3\nexp{11}$
protons \cite{gschwendtner:2016}. Only half of these are driving the wakefield, presuming the self-modulation is seeded
in the middle of the unmodulated bunch. In addition, a significant portion of the protons are lost from the near axis
region during self-modulation. The needed gap between the two plasma cells also contributes to a decreased density of
protons near the axis, and some of the field is drained by other protons seeing an accelerating phase within the drive
beam itself. In total, the expected peak $E_{z}$ field of $~2\funit{GV}{m}$ in the first plasma cell drops to
$500-600\funit{MV}{m}$ in the second cell \cite{awake_collaboration:2016}. A single proton bunch of $1.46\nexp{10}$,
or $2.34\unit{nC}$ and $7\unit{kA}$ was able to drive an $E_{z}$ field of a comparable magnitude in our single drive
bunch setup.

As the baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble, the
plasma electrons are only depleted to around $65\%$ of nominal plasma density at the point where we inject the electron
bunch in reference simulations \cite{awake_collaboration:2016}. This condition is replicated in out single bunch case.
The peak density of the drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch setup uses
the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size $\sigma_{r} = 200\unit{\mu m}$
throughout the plasma cell. The length of the drive bunch in these simulations is $\sigma_{z} = 40\unit{\mu m}$.
In order to create a stable environment to study the evolution of the witness bunch we prevented the proton beam from
evolving radially by increasing the particle mass by a factor of $1\mexp{6}$. We are not considering the evolution of
the proton beam in this study.

@@

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}, and the number of inter-dependent parameters is already large for these simulations. AWAKE
baseline energy for the long witness beam for Run 1 is $W_{eb} = 10-20\unit{MeV}$ \cite{gschwendtner:2016}, but in Run 2
the experiment will likely require a higher initial energy A $W_{eb} \approx 50\unit{MeV}$ is sufficient to minimise
de-phasing as the drift $\Delta\xi \propto \gamma^{-2}$ \cite{berglyd_olsen:2015}.

Earlier simulations were done using Run 1 witness bunch parameters, with radius $\sigma_{r} = 100\unit{\mu m}$. This
caused the bunch to undergo rapid transverse evolution as the it entered the plasma region, reducing its $\sigma_{r}$
from $100\unit{\mu m}$ to $< 10\unit{\mu m}$ \cite{berglyd_olsen:2016}. This transverse evolution was primarily due to a
mismatch between beam emittance and the plasma density, but was difficult to avoid as narrow on-axis bunches were
numerically difficult to resolve sufficiently in cylindrical geometry.

In these 3D simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{r}$ of $5.25\unit{\mu m}$. The relatively small size of the witness beam compared to the proton beam put some
restrictions on the transverse resolution of the witness beam also in these simulations. The transverse grid cell
resolution was set to $1.17\unit{\mu m}$, and the witness bunch was simulated with $16.8+unit{M}$ non-weighted
particles.
